Reasoning Graphically about Quantum Computation

### 2008-03-05 papers/reasoning-graphically/paper.tex
more work.
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1402 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\begin{abstract}
  Recent graph-based formalisms of quantum computation provide an
  abstract and symbolic way to represent and simulate the
  computations. However, manual manipulation of such graphs is slow
  and error prone. We present a formalism that supports mechanised
  reasoning about such graphs. This involves a compositional account
  of graph rewriting that preserves the underlying categorical
  semantics. We present a generic system with a fixed logical kernel
  for representing compact closed categories. This provides a
  declarative account which supports derived rules. We illustrate the
  framework by instantiating it for the graphical language of quantum
  computation.

  \keywords{graph rewriting, quantum computing, categorical logic,
    interactive theorem proving, graphical calculi; concerning:
    symbolic computation, automated reasoning, formal mathematics,
    logic}
\end{abstract}

@@

The major problem with these graphical representations is the lack of
machinery for automating the manipulation of graphs.  Existing
approaches to graph transformation have a different underlying
semantics corresponding to graphs in general rather than those which
describe compact closed categories. The result is that they provide an
unsound concept of rewriting for the kind of graphs we are interested
in.

This paper reviews a formalism for graphical representations of
compact closed categories. We also revisit a model of quantum
computation based on this calculus. We then extend the graphical
calculus in two significant ways driven by the need to express rules
that could not be accounted for in the initial formalism. We provide a
semantics for these in terms of the initial representation. By
suitably combining these extensions to graphs, we provide a
representation that captures an interesting and useful set of graph
patterns. We use this as the foundation for a simple logical framework
for manipulating compact closed categories as graphs. This has a
suitable rewriting mechanism where the axioms of the underlying
object-formalism are expressed as equations between graphs. We then
present a short case study that illustrates the framework by
instantiating it for the introduced model of quantum computation.
This shows how the framework can be used to symbolically perform
simplifications to quantum programs as well as simulate quantum
computations.

@@

\noindent These ensure that the structure of the graph is preserved by
the morphism: an edge connected to a node gets mapped to a new edge
that must connected in the same way to the mapped node.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category is then given by all graphs formed by
composing and tensoring these basic graphs.  Note  that since $Q$ is
self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.

@@

A crucial feature of rules like the Spider Theorem is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing (pronounced band-boxing), on
graph representations. Given a graph representation, this introduces a
new notation by outlining a set of nodes. These nodes are said to be
in a !-box. Intuitively, the resulting !-boxed graph can be thought of
as representing a set of graphs with an arbirary number of copies of
the !-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies. This erases all edges to and from the !-box.

@@

\subsection{!-Matching.}

@@

\subsection{!-Box Semantics.}

@@

\subsection{Matching}

The purpose of graph patterns is to formally express, in a declarative
manner, the intuitive idea of graph matching which commonly used in
quantum computation. To do this, we combine the two extensions to the
concrete graph representation, namely variable nodes and !-boxes, to
provides a matching relation between graph patterns and concrete
graphs as well as directly between graph patterns. The former relation
gives a semantics for our system and ensures soundness, while the
latter supports deirved equations and thus provides the theoretical
basis for our formulation of rewriting.

\subsubsection{Concrete Matches.}

The semantics for a graph pattern $G$ is a set of concrete graphs
denoted by $\minterp{G}$. We say that members of this set are
\emph{matched} by $G$. Formally, the concrete set of matches are
defined as:

@@

@@

The specification for one pattern matching another one is simply that
it is more general. Formally, a graph pattern, $G$, matches another
graph pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This is
the specification for a matching algorithm which we implement of
in~\S\ref{sec:rewriting}. We can use the definition of open embedding
from variable-node graphs as an intermediate mechanism and show that a
suitable definition of graph pattern matching is simply the !-Box
matching followed by variable-node matching:

\begin{definition}
  $G \leq H \Leftrightarrow \forall H' \in \binterp{H}.\,\exists G' \in \binterp{G}.\, G' \leq_v H'$
\end{definition}

Because variable-node matching preserves the semantics, it is trivial to
prove that this definition meets the specification.

@@

In this section we describe how the garph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the specification
provided by the underlying model an \emph{object-level} graph
formalism. In particular, an object-level formalisation defines a set
of rules which are treated as the axioms for the system. It also
defines the data at the nodes and edges as well as corresponding
data-matching behaviour. For it's part, the meta-level provides
machinery to manipulate graphs and derive new rules. In this section
we describe the meta-level framework and prove its adequacy for
rewriting.

@@

\subsubsection{Adequacy.} An equation on graph patterns corresponds to
a possibly infinite set of equations betweens concrete graphs. We
might expect that such equations express too much: they may allow
rewrites which are not intended. We prove \emph{adequacy}, which
states that this is not the case. In particular, given an equation
between concrete graphs, $G = H$, the equations between the lifting of
these to variable node graphs by replacing boundary nodes with
variable nodes is a conservative extension. The proof follows directly
from the obervation that every instance of the lifted variable node
equation corresponds to compsition of $G$ and $H$ with some other
graph (the non-embedded part).

@@

\begin{center}
\prooftree
A = B \in \Gamma
\justifies
\Gamma \vdash A = B
\using\mbox{trivial}
\endprooftree
\quad\quad
\prooftree
\justifies
\Gamma \vdash A = A
\using\mbox{reflexivity}
\endprooftree
\end{center}

\begin{center}
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{symmetry}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{substitution}
\endprooftree
\end{center}

@@

This system allows derived rules which allow a combination of steps to
be abbreviated by representing them as a single new rule.

@@

\section{Implementation}
\label{ssec:implementation}

@@

\subsection{}
\label{ssec:node-expressions}

>>>>

\begin{abstract}
  Recent graph-based formalisms of quantum computation provide an
  abstract and symbolic way to represent and simulate
  computations. However, manual manipulation of such graphs is slow
  and error prone. We present a formalism, based on compact closed
  categories, that supports mechanised reasoning about such
  graphs. This gives a compositional account of graph rewriting that
  preserves the underlying categorical semantics. Using this
  representation, we describe a generic system with a fixed logical
  kernel that supports reasoning about models of compact closed
  category. A salient feature of the system is that it provides a
  declarative account of derived results. We illustrate the framework
  by instantiating it for a graphical language of quantum computation
  and show how this can be used to perform symbolic computation.

  \keywords{graph rewriting, quantum computing, categorical logic,
    interactive theorem proving, graphical calculi; concerning:
    symbolic computation, automated reasoning, formal mathematics}
\end{abstract}

@@

The major problem with these graphical representations is the lack of
machinery for automating their manipulation. Existing approaches to
graph transformation have a different underlying semantics
corresponding to graphs in general rather than those which describe
compact closed categories. The result is that they provide an unsound
concept of rewriting for the kind of graphs we are interested in.

The main contribution of this paper is a graph-vased formalism that is
suitable for representing and evaluating quantum computations. We
start by reviewing a formalism for graphical representations of
compact closed categories. We also revisit a model of quantum
computation based on this calculus. We then extend the graphical
calculus in two significant ways driven by the need to express rules
that could not be accounted for in the initial formalism. By combining
these extensions, we provide a representation that captures an
interesting and useful set of graph patterns. Notably, it can express
the Spider Theorem, which is normally written using informal elipses
notation. We provide a semantics for our graph-based formalism in
terms of the initial representation of compact closed categories.

Using our graph-based formalism as the representational foundation, we
develop a simple logical framework for manipulating models of compact
closed categories. This has a suitable rewriting mechanism where the
axioms of the underlying object-formalism are expressed as equations
between graphs. We then present a short case study that illustrates
the framework by instantiating it for the introduced model of quantum
computation.  This shows how the framework can be used to symbolically
perform simplifications of quantum programs as well as simulate
computations.

@@

\noindent These ensure that the structure of the graph is preserved by
the morphism: an edge connected to a node gets mapped to a new edge
that must be connected, in the same way, to the mapped node.

@@

Coecke and Duncan \cite{Coecke:2008jo} propose a formal algebraic
system for quantum computation built from the following collection of
generators:  
\begin{description}
\item[Objects:] the self dual object $Q$;
\item[Arrows:] two families of arrows $X$ and $Z$:
  \begin{align*}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align*}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
To each each arrow $f : A\to B$ we assign a formal adjoint $f^\dag : B
\to A$.  Each arrow is represented as a small graph; its adjoint is
the same graph written upside down.  We use colours to denote two families:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  (The adjoints
for $H$, $\alpha_X$ and $\alpha_Z$ will be defined equationally.)  The
free compact closed category is then given by all graphs formed by
composing and tensoring these basic graphs.  Note  that since $Q$ is
self dual, that is $Q = Q^*$, we use \emph{undirected} graphs.

@@

A crucial feature of rules like the Spider Theorem is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing (pronounced bang-boxing), on
graph representations. Given a graph representation, this introduces a
new notation by outlining a set of nodes. These nodes are said to be
in a !-box. Intuitively, the resulting !-boxed graph can be thought of
as representing a set of graphs with an arbirary number of copies of
the !-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies. This erases all edges to and from the !-box.  The Spider
Theorem in Figure~\ref{fig:informal-spider-law} for instance is
represented with !-boxes in Figure~\ref{fig:spider-thm-patt}.

@@

\subsubsection{Matching.}

@@

\subsubsection{Semantics.}

@@

% \subsubsection{Matching.}

% The purpose of graph patterns is to formally express, in a declarative
% manner, the intuitive idea of graph matching which commonly used in
% quantum computation. To do this, we combine the two extensions to the
% concrete graph representation, namely variable nodes and !-boxes, to
% provides a matching relation between graph patterns and concrete
% graphs as well as directly between graph patterns. The former relation
% gives a semantics for our system and ensures soundness, while the
% latter supports deirved equations and thus provides the theoretical
% basis for our formulation of rewriting.

\subsubsection{Semantics as Concrete Matches.}

The semantics for a graph pattern $G$ is a set of concrete graphs
denoted by $\minterp{G}$ and define it as:

@@

\noindent this simply considers every interpretation of the !-boxes to
give variable-node graphs for which we then appeal to their own
semantics.

@@

The specification for one pattern, $G$, to match another one, $H$ is
that it is more general with respect to the interpretation:
$\minterp{G} \supseteq \minterp{H}$. However, graph patterns can
correspond to a countably infinite number of concrete graphs. Thus
matching between graph pattern cannot be implemented by simply
unfolding them to their interpretated concrete graphs and checking the
membership relation.

*** add matching algo outline ***

%We can use the definition of open embedding from variable-node graphs
%as an intermediate mechanism and show that a suitable definition of
%graph pattern matching is simply the !-Box matching followed by
%variable-node matching:
%
%\begin{definition}
%  $G \leq H \Leftrightarrow \forall H' \in \binterp{H}.\,\exists G' \in \binterp{G}.\, G' \leq_v H'$
%\end{definition}
%Because variable-node matching preserves the semantics, it is trivial to
%prove that this definition meets the specification.

@@

In this section we describe how the garph pattern formalism can
provide a \emph{meta-level} framework for reasoning about models of
compact closed categories. Following the terminology of logical
frameworks such as Isabelle~\cite{isabelle}, we call the
specification, provided by the underlying model, an
\emph{object-level} graph formalism. An object-level formalisation
defines a set of rules which are treated as the axioms for the
system. It also defines the data at the nodes and edges as well as
corresponding data-matching behaviour. For it's part, the meta-level
provides generic machinery to manipulate graphs and derive new
rules. In this section we describe the meta-level framework and prove
its adequacy for rewriting.

@@

\subsection{Lifting}

*** lifting concrete laws to graph-pattern laws ***

\subsection{Adequacy}

An equation on graph patterns corresponds to
a possibly infinite set of equations betweens concrete graphs. We
might expect that such equations express too much: they may allow
rewrites which are not intended. We prove \emph{adequacy}, which
states that this is not the case. In particular, given an equation
between concrete graphs, $G = H$, the equations between the lifting of
these to variable node graphs by replacing boundary nodes with
variable nodes is a conservative extension. The proof follows directly
from the obervation that every instance of the lifted variable node
equation corresponds to compsition of $G$ and $H$ with some other
graph (the non-embedded part).

@@

\begin{center}
%% not needed, can be derived by refl + subst
%\prooftree
%A = B \in \Gamma
%\justifies
%\Gamma \vdash A = B
%\using\mbox{trivial}
%\endprooftree
%\quad\quad
\prooftree
\justifies
\Gamma \vdash A = A
\using\mbox{refl}
\endprooftree 
\quad\quad
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{sym}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{subst}
\endprooftree
\end{center}

@@

A new graph can be introduced using the reflexivity law. This can be
the left hand side of a conjecture, or a graph that needs to be
computed. By applying the \emph{subst} rule, intermediate results are
derived which can themselves be used to rewrite other rules and
conjetcures. In this way, the system allows derived rules to provide
an abbreviation for a combination of steps.

@@

%\subsection{Implementation}
%\label{ssec:implementation}

@@
